## Q2

---
title: "Q2. Exploratory Analysis"
author: "Salman"
date: "2024-04-05"
output: word_document
---
```{r}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)
library(ggplot2)

penguins2 <- as.data.frame(palmerpenguins::penguins)
penguins2 <- penguins2[, -c(2, 7, 8)]  
penguins2 <- na.omit(penguins2)         

summary(penguins2)

aggregate(. ~ species, data = penguins2, mean)

ggplot(penguins2, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point() +
  labs(title = "Bill Length vs. Bill Depth by Species", 
       x = "Bill Length (mm)", y = "Bill Depth (mm)")

ggplot(penguins2, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot() +
  labs(title = "Flipper Length by Species", x = "Species", y = "Flipper Length (mm)")

ggplot(penguins2, aes(x = body_mass_g, fill = species)) +
  geom_density(alpha = 0.7) +
  labs(title = "Body Mass Distribution by Species", x = "Body Mass (g)")


## Q3

---
title: "Q3. Splitting the Dataset"
author: "Salman"
date: "2024-04-05"
output: word_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)
library(caret)

penguins2 <- as.data.frame(palmerpenguins::penguins)
penguins2 <- penguins2[, -c(2, 7, 8)]  
penguins2 <- na.omit(penguins2)         

set.seed(123)
training_indices <- createDataPartition(y = penguins2$species, p = 0.7, list = FALSE)

# Splitting the dataset into training and testing sets
training_set <- penguins2[training_indices, ]
testing_set <- penguins2[-training_indices, ]

# Output the size of the training and testing sets
cat("Training set rows:", nrow(training_set), "\n")
cat("Testing set rows:", nrow(testing_set), "\n")


## Q4

---
title: "Q4. k Nearest Neighbor Analysis"
author: "Salman"
date: "2024-04-05"
output: word_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(class)
library(mclust)
library(palmerpenguins)


data("penguins", package = "palmerpenguins")
penguins2 <- as.data.frame(penguins)


penguins2 <- penguins2[, -c(2, 7, 8)] 
penguins2 <- na.omit(penguins2)

set.seed(123)
training_indices <- createDataPartition(y = penguins2$species, p = 0.7, list = FALSE)
training_set <- penguins2[training_indices, ]
testing_set <- penguins2[-training_indices, ]

train_control <- trainControl(method="cv", number=10) 

# Train the k-NN model
set.seed(123) 
knn_model <- train(species ~ ., data=training_set, method="knn", tuneLength=20, trControl=train_control)

# Print 
print(knn_model$bestTune)

# Predictions on the testing set
knn_predictions <- predict(knn_model, newdata=testing_set)

# Confusion matrix
confusion_matrix <- confusionMatrix(knn_predictions, testing_set$species)
print(confusion_matrix)

# Misclassification rate
misclassification_rate <- 1 - sum(diag(confusion_matrix$table)) / sum(confusion_matrix$table)
cat("Misclassification Rate:", misclassification_rate, "\n")

# Adjusted Rand Index (ARI)
ARI <- adjustedRandIndex(knn_predictions, testing_set$species)
cat("Adjusted Rand Index (ARI):", ARI, "\n")


## Q5

---
title: "Q5. Classification Tree"
author: "Salman"
date: "2024-04-05"
output: word_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(rpart)
library(rpart.plot)
library(mclust)
library(palmerpenguins)


data("penguins", package = "palmerpenguins")
penguins2 <- as.data.frame(penguins)


penguins2 <- penguins2[, -c(2, 7, 8)]
penguins2 <- na.omit(penguins2)

set.seed(123)
training_indices <- createDataPartition(y = penguins2$species, p = 0.7, list = FALSE)
training_set <- penguins2[training_indices, ]
testing_set <- penguins2[-training_indices, ]

# Fit the classification tree
tree_model <- rpart(species ~ ., data=training_set, method="class")

# Plot the classification tree
rpart.plot(tree_model, main="Classification Tree for Penguin Species")

# Predict on the testing set
tree_predictions <- predict(tree_model, newdata=testing_set, type="class")

# Confusion matrix
confusion_matrix_tree <- confusionMatrix(tree_predictions, testing_set$species)
print(confusion_matrix_tree)

# Misclassification rate
misclassification_rate_tree <- 1 - sum(diag(confusion_matrix_tree$table)) / sum(confusion_matrix_tree$table)
cat("Misclassification Rate:", misclassification_rate_tree, "\n")

# Adjusted Rand Index (ARI)
ARI_tree <- adjustedRandIndex(tree_predictions, testing_set$species)
cat("Adjusted Rand Index (ARI):", ARI_tree, "\n")


## Q6

---
title: "Q6. Random Forest Analysis"
author: "Salman"
date: "2024-04-05"
output: word_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
library(randomForest)
library(caret)
library(mclust)
library(palmerpenguins)

penguins2 <- as.data.frame(palmerpenguins::penguins)


penguins2 <- penguins2[, -c(2, 7, 8)] 
penguins2 <- na.omit(penguins2)

set.seed(123)
training_indices <- createDataPartition(y = penguins2$species, p = 0.7, list = FALSE)
training_set <- penguins2[training_indices, ]
testing_set <- penguins2[-training_indices, ]

train_control <- trainControl(method="cv", number=5, search="grid")
tune_grid <- expand.grid(mtry = seq(2, sqrt(ncol(training_set)-1), by = 1))

# Tuning of the model
set.seed(123)
rf_tune_result <- train(species ~ ., data=training_set, method="rf", trControl=train_control, tuneGrid=tune_grid, ntree=500)
print(rf_tune_result$bestTune)

# Tuned random forest model
best_mtry <- rf_tune_result$bestTune$mtry
set.seed(123)
tuned_rf <- randomForest(species ~ ., data=training_set, mtry=best_mtry, ntree=500)

# Prediction and evaluation
rf_predictions <- predict(tuned_rf, newdata=testing_set)
confusion_matrix_rf <- confusionMatrix(rf_predictions, testing_set$species)
print(confusion_matrix_rf)

# Misclassification rate
misclassification_rate_rf <- 1 - sum(diag(confusion_matrix_rf$table)) / sum(confusion_matrix_rf$table)
cat("Misclassification Rate:", misclassification_rate_rf, "\n")

# Adjusted Rand Index (ARI)
ARI_rf <- adjustedRandIndex(rf_predictions, testing_set$species)
cat("Adjusted Rand Index (ARI):", ARI_rf, "\n")



